\documentclass{article}

\usepackage{fullpage}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{graphicx}
\usepackage{xcolor}
\usepackage{hyperref}
\usepackage{tikz}

\newcounter{placeholderCount}
\newcommand{\placeholder}[1]{{
    \stepcounter{placeholderCount}
    \color{red} [#1]
}}

\newcounter{solutionCount}
\newcommand{\solution}[1]{
    \stepcounter{solutionCount}
    \paragraph{Solution \thesolutionCount.}
    {
        #1
    }
}

\title{
    Deep Learning (Spring 2026) \\
    Solution for Assignment 5
}
\author{
    Pan Changxun
    \\
    2024011323
}

\begin{document}

\maketitle

%% Problem 1 (Dequantization bound, 30 pts)
\solution{
    We want to show $\mathcal{L}(\theta) \leq \mathbb{E}_{x \sim p_{\mathrm{data}}(x)}[\log p(x)]$.

    By definition:
    \[
      \mathcal{L}(\theta) = \mathbb{E}_{x \sim p_{\mathrm{data}}} \mathbb{E}_{u \sim \mathrm{Uniform}[0,1)^d} [\log \hat{p}(x + u)].
    \]

    Since $p(x) = \int_{[0,1)^d} \hat{p}(x + u) \, du$, and the uniform distribution over $[0,1)^d$ has density $1$, we can write:
    \[
      p(x) = \mathbb{E}_{u \sim \mathrm{Uniform}[0,1)^d}[\hat{p}(x + u)].
    \]

    By Jensen's inequality, since $\log$ is a concave function:
    \[
      \mathbb{E}_{u}[\log \hat{p}(x + u)] \leq \log \mathbb{E}_{u}[\hat{p}(x + u)] = \log p(x).
    \]

    Taking expectation over $x \sim p_{\mathrm{data}}(x)$ on both sides:
    \[
      \mathbb{E}_{x \sim p_{\mathrm{data}}} \mathbb{E}_{u}[\log \hat{p}(x + u)] \leq \mathbb{E}_{x \sim p_{\mathrm{data}}}[\log p(x)].
    \]

    The left-hand side is exactly $\mathcal{L}(\theta)$, so:
    \[
      \mathcal{L}(\theta) \leq \mathbb{E}_{x \sim p_{\mathrm{data}}(x)}[\log p(x)].
    \]
    $\hfill\blacksquare$
}

%% Problem 2 (Autoregressive Flow, 30 pts)
\solution{
    To show that $T$ is a valid normalizing flow, we need to verify two properties:
    \begin{enumerate}
        \item $T$ is invertible (i.e., there exists a bijection between $u$ and $x$).
        \item The Jacobian determinant of $T$ can be computed efficiently.
    \end{enumerate}

    \textbf{(1) Invertibility.}

    The forward transformation is defined recursively as:
    \[
      x_i = \mu_i(x_{<i}) + \exp(\alpha_i(x_{<i})) \cdot u_i, \quad i = 1, 2, \ldots, L.
    \]

    We can invert this to recover $u$ from $x$:
    \[
      u_i = \frac{x_i - \mu_i(x_{<i})}{\exp(\alpha_i(x_{<i}))}, \quad i = 1, 2, \ldots, L.
    \]

    Since $\exp(\alpha_i(x_{<i})) > 0$ for all $i$, each $u_i$ is uniquely determined given $x_1, \ldots, x_i$. The inverse can be computed sequentially: first $u_1$ from $x_1$, then $u_2$ from $x_1, x_2$, and so on. Therefore, $T$ is a bijection.

    \textbf{(2) Tractable Jacobian determinant.}

    The Jacobian matrix $J = \frac{\partial x}{\partial u}$ has entries:
    \[
      J_{ij} = \frac{\partial x_i}{\partial u_j}.
    \]

    Since $x_i = \mu_i(x_{<i}) + \exp(\alpha_i(x_{<i})) \cdot u_i$ and $x_i$ depends on $u_1, \ldots, u_{i-1}$ only through $x_1, \ldots, x_{i-1}$ (which are computed from $u_1, \ldots, u_{i-1}$), and directly on $u_i$:
    \begin{itemize}
        \item For $j > i$: $x_i$ does not depend on $u_j$, so $J_{ij} = 0$.
        \item For $j = i$: $\frac{\partial x_i}{\partial u_i} = \exp(\alpha_i(x_{<i}))$.
    \end{itemize}

    Thus, $J$ is a \textbf{lower triangular} matrix with diagonal entries $\exp(\alpha_i(x_{<i}))$. The determinant of a triangular matrix is the product of its diagonal entries:
    \[
      \det(J) = \prod_{i=1}^{L} \exp(\alpha_i(x_{<i})) = \exp\left(\sum_{i=1}^{L} \alpha_i(x_{<i})\right).
    \]

    The log-absolute-determinant is therefore:
    \[
      \log |\det(J)| = \sum_{i=1}^{L} \alpha_i(x_{<i}),
    \]
    which can be computed in $O(L)$ time.

    \textbf{(3) Resulting density.}

    By the change-of-variables formula, the density of $x = T(u)$ is:
    \begin{align}
      p(x) &= p_U(T^{-1}(x)) \cdot |\det(J^{-1})| \nonumber \\
      &= p_U(u) \cdot \prod_{i=1}^{L} \frac{1}{\exp(\alpha_i(x_{<i}))} \nonumber \\
      &= \prod_{i=1}^{L} \mathcal{N}(u_i \mid 0, 1) \cdot \frac{1}{\exp(\alpha_i(x_{<i}))} \nonumber \\
      &= \prod_{i=1}^{L} \frac{1}{\exp(\alpha_i(x_{<i}))} \cdot \frac{1}{\sqrt{2\pi}} \exp\left(-\frac{u_i^2}{2}\right). \nonumber
    \end{align}

    Substituting $u_i = \frac{x_i - \mu_i(x_{<i})}{\exp(\alpha_i(x_{<i}))}$:
    \begin{align}
      p(x) &= \prod_{i=1}^{L} \frac{1}{\exp(\alpha_i(x_{<i})) \sqrt{2\pi}} \exp\left(-\frac{(x_i - \mu_i(x_{<i}))^2}{2\exp(\alpha_i(x_{<i}))^2}\right) \nonumber \\
      &= \prod_{i=1}^{L} \mathcal{N}\left(x_i \mid \mu_i(x_{<i}), \exp(\alpha_i(x_{<i}))^2\right). \nonumber
    \end{align}

    This recovers exactly the autoregressive factorization $p(x) = \prod_{i=1}^{L} p(x_i | x_{<i})$ with Gaussian conditionals, confirming that $T$ is a valid normalizing flow. $\hfill\blacksquare$
}

%% Problem 3 (PixelCNN, 30 pts)
\solution{
    \begin{enumerate}
        \item \textbf{Sawtooth-shaped receptive fields from masked convolutions.}

        We use the $5 \times 5$ masked kernel from Figure~2 as a concrete example, with half-width $h = \lfloor 5/2 \rfloor = 2$. The mask $m$ is:
        \[
        m = \begin{pmatrix}
        1 & 1 & 1 & 1 & 1 \\
        1 & 1 & 1 & 1 & 1 \\
        1 & 1 & 0 & 0 & 0 \\
        0 & 0 & 0 & 0 & 0 \\
        0 & 0 & 0 & 0 & 0
        \end{pmatrix}.
        \]

        This mask has two distinct regions:
        \begin{itemize}
            \item \emph{Upper rows} (rows above center): all 5 columns are 1 --- the kernel sees the \textbf{full width}.
            \item \emph{Center row}: only the 2 columns to the left of center are 1 --- the kernel only sees \textbf{left neighbors}.
            \item \emph{Lower rows}: all 0 --- completely invisible.
        \end{itemize}

        \textbf{After one layer} ($n=1$): The receptive field (RF) of the output pixel at position $(r, c)$ is:
        \begin{itemize}
            \item Rows $r{-}2$ and $r{-}1$: columns $[c{-}2, \, c{+}2]$ (full width, 5 pixels each).
            \item Row $r$: columns $[c{-}2, \, c{-}1]$ (only 2 pixels to the left).
        \end{itemize}
        The right boundary jumps from $c{+}2$ (at row $r{-}1$) to $c{-}1$ (at row $r$), an abrupt step of 3 columns.

        \textbf{After two layers} ($n=2$): Each position in layer~2 applies the same masked kernel to the layer-1 output. The RF of $(r,c)$ in the original input is the union of 1-layer RFs of all positions that layer~2 reads. The result is:
        \begin{itemize}
            \item Rows $r{-}4$ and $r{-}3$: right boundary at $c{+}4$ (full symmetric growth).
            \item Rows $r{-}2$ and $r{-}1$: right boundary at approximately $c{+}2$ (reduced rightward extent).
            \item Row $r$: right boundary at $c{-}3$ (only leftward growth on the current row).
        \end{itemize}

        \textbf{General pattern after $n$ layers:} The asymmetry accumulates with depth. The right boundary of the RF exhibits a \textbf{sawtooth} (staircase) shape:
        \begin{itemize}
            \item On \emph{upper rows} of the mask ($h = 2$ rows at a time for a $5\times 5$ kernel), the right boundary remains \emph{flat} (same rightward extent) for $h$ consecutive rows, forming a ``tooth.''
            \item Then it \emph{drops} by $h$ columns as we move down to the next group of rows.
            \item On the \emph{current row}, the right boundary is far to the left of center, since information on the same row can only propagate leftward.
        \end{itemize}

        This periodic pattern of ``flat plateau $\to$ step down $\to$ flat plateau $\to$ step down'' is the characteristic sawtooth shape shown in Figure~1a. The upper-right region that is never covered by the RF constitutes the \textbf{blind spot}.

        \item \textbf{Removing the blind spot.}

        The blind spot arises because the masked convolution on the current row can only see pixels to the left, preventing information from the upper-right region from reaching the current pixel through the same-row path.

        The solution, proposed by Gated PixelCNN (van den Oord et al., 2016), is to split the computation into two streams: a \textbf{vertical stack} and a \textbf{horizontal stack}.

        Denote:
        \begin{itemize}
            \item $x$: input feature map.
            \item $w_v$: vertical convolution kernel (unmasked in full rows above the center row, including the center row itself; all columns below center row are zero).
            \item $m_v$: vertical mask -- ones for all rows up to and including the center row, zeros below.
            \item $w_h$: horizontal convolution kernel ($1 \times k$ kernel, masked to only include the current position and positions to its left).
            \item $m_h$: horizontal mask -- a $1 \times k$ mask with ones for columns left of and including center.
            \item $W_{vh}$: a $1 \times 1$ convolution that feeds information from the vertical stack to the horizontal stack.
        \end{itemize}

        The per-layer computation is:

        \textbf{Step 1 -- Vertical stack:}
        \[
            v = \mathrm{conv}(m_v \cdot w_v, \; x)
        \]
        This uses a vertically-masked kernel that sees all pixels in rows above and including the center row (across the full width). The output is then \textbf{shifted down by one row} so that position $(r,c)$ in the vertical stack output only contains information from rows strictly above $r$, preserving the autoregressive property.

        \textbf{Step 2 -- Horizontal stack with vertical input:}
        \[
            h = \mathrm{conv}(m_h \cdot w_h, \; x) + W_{vh} \cdot v
        \]
        The horizontal stack uses a $1 \times k$ masked convolution that only sees pixels to the left on the same row. It receives a $1 \times 1$ convolution of the (shifted) vertical stack output as additional input, which carries full-width information from all rows above.

        \textbf{Why this removes the blind spot:}
        \begin{itemize}
            \item The vertical stack sees all columns in rows above the current pixel (no horizontal masking restriction in upper rows), so the upper-right region (the former ``blind spot'') is fully captured.
            \item The vertical-to-horizontal connection ($W_{vh} \cdot v$) feeds this information into the horizontal stack, which also has left-context from the current row.
            \item The downward shift of the vertical stack output ensures that the pixel at position $(r,c)$ never conditions on itself or any pixel below/right of it, maintaining the autoregressive ordering.
        \end{itemize}

        Optionally, residual connections can be added to the horizontal stack:
        \[
            h_{\mathrm{out}} = h + x_h
        \]
        where $x_h$ is the horizontal stack input from the previous layer.
    \end{enumerate}
}

%% Problem 4 (AI Usage Disclosure, 10 pts)
\solution{
    Used Claude (AI assistant by Anthropic) in the following capacities:
    \begin{itemize}
        \item \textbf{Solution ideation:} Brainstorming approaches and generating initial ideas for problem-solving strategies, particularly for structuring mathematical proofs and derivations.
        \item \textbf{English grammar and writing:} Polishing and refining the English writing throughout the solutions to improve clarity, coherence, and academic tone.
        \item \textbf{\LaTeX{} formatting:} Optimizing typesetting and layout of mathematical expressions, tables, and document structure for better readability and presentation.
        \item \textbf{Result verification:} Cross-checking the correctness of intermediate computations, final answers, and logical consistency of proofs.
    \end{itemize}
    All final solutions were reviewed and verified by the author.
}

\ifnum\value{placeholderCount}>0
    \vfill
    \begin{center}
        \color{red} \framebox{\textbf{WARNING: PLEASE REMOVE ALL PLACEHOLDERS BEFORE SUBMISSION}}
    \end{center}
\fi

\end{document}
